---
title: "export_data"
author: "Group P"
date: '2022-09-27'
output: pdf_document
---

This is a team assignment; each team should complete and turn in a PDF created from an Rmd via Github. Please include code and output for the following components:

Description of data set
1. What is the data source? (1-2 sentences on where the data is coming from, dates included, etc.)
2. How does the data set relate to the group problem statement and question?



\newpage 
Import statement
NOTE: Please use data sets available in the PHW251 Project Data github repo Links to an external site. (this is important to make sure everyone is using the same data sets)(done)
Use appropriate import function and package based on the type of file(done)
Utilize function arguments to control relevant components (i.e. change column types, column names, missing values, etc.)(working on it)
Document the import process(working on it)

Loading the library to be used in this assignment
```{r}
library(tidyverse)
library(readr)
library(janitor)
```

Importing data from git repositories
```{r}
file_path1<-"https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_mpx_cases.csv"
monkey_pox <-read_csv(file_path1)%>% clean_names()

file_path2 <-"https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_pop_denominators.csv" 
pop_denominator<- read.csv(file_path2)%>% clean_names()

file_path3 <- "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_census_stats.csv"
census_stats <- read.csv(file_path3)%>% clean_names()

file_path4<- "https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/world_country_regions.csv"

world_country_region <- read.csv(file_path4)%>% clean_names()
```


\newpage
Data manipulation and cleaning
```{r}
#aggregating the monkey pox data set to the desired time period (week or month) 
monkey_pox

library(lubridate)
 monkey_pox <- monkey_pox%>% mutate(months_date= month(date_rep))
 monkey_pox
```
Joining monkey pox data set with pop denominator 

```{r}
print(pop_denominator)
merged_data <- merge(new_az_players_dataset, new_pregame_stats, by="player")
```
\newpage

Identify data types for 5+ data elements/columns/variables
Identify 5+ data elements required for your specified scenario. If <5 elements are required to complete the analysis, please choose additional variables of interest in the data set to explore in this milestone.
Utilize functions or resources in RStudio to determine the types of each data element (i.e. character, numeric, factor)
Identify the desired type/format for each variable—will you need to convert any columns to numeric or another type? 

\newpage
Provide a basic description of the 5+ data elements
Numeric: mean, median, range
Character:  unique values/categories
Or any other descriptives that will be useful to the analysis