---
title: "Milestone_5"
author: "Moliehi Mokete and Bongekile Nkosi"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Problem statement**

With the rise of MonkeyPox cases in Europe, in preparation of an appropriate
response, our team has been tasked with developing an update for the leadership 
of our state health department regarding the state of MPX. We are specifically 
interested in understanding how case rates may differ by region and various 
demographic factors. The demographic factor we are particularly interested in
is gender.

```{r, include=FALSE,echo=FALSE}
library(tidyverse)
library(readr)
library(janitor)
library(purrr)
library(stringr)
library(kableExtra)
library(plotly)
library(DT)
```


```{r,include=FALSE, echo=FALSE}
file_path1<-"https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_mpx_cases.csv"
monkey_pox <-read_csv(file_path1,na=c("","NA","*", "n/a"),
                      show_col_types = FALSE)%>% clean_names()

monkey_pox<- monkey_pox%>% mutate(month_rep= months(date_rep))%>%
  group_by(country_code,month_rep)%>%
  summarise(total_conf_case = sum(conf_cases))



 var_info_MPX <- data.frame(Variable = names(monkey_pox),
                      VariableType = sapply(monkey_pox, class),
                      MissingValues = sapply(monkey_pox, function(y) 
                        sum(length(which(is.na(y))))),
                      row.names = NULL)
 var_info_MPX 
 
 
 monkey_pox<-monkey_pox %>% 
   select(country_code,month_rep,total_conf_case)

```




```{r,include=FALSE,echo=FALSE}
file_path2<-"https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_pop_denominators.csv" 
pop_denominator<- read.csv(file_path2,na = c("", "NA", "*", "n/a")) %>%
  clean_names() %>% 
  rename(country_code = geo)

 var_info_PD <- data.frame(Variable = names(pop_denominator),
                      VariableType = sapply(pop_denominator, class),
                      MissingValues = sapply(pop_denominator, function(y) 
                        sum(length(which(is.na(y))))),
                      row.names = NULL)
 var_info_PD 
 
 pop_denominator <- pop_denominator%>% 
   filter(time_period== 2022)%>%
   select(country_code, time_period)

```


```{r,include=FALSE,echo=FALSE}
file_path4<-"https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/world_country_regions.csv"
world_country_region <- read.csv(file_path4,na = c("", "NA", "*", "n/a"))%>%
  clean_names()

var_info_WCR <- data.frame(Variable = names(world_country_region),
                      VariableType = sapply(world_country_region, class),
                      MissingValues = sapply(world_country_region, function(y) 
                        sum(length(which(is.na(y))))),
                      row.names = NULL)
 var_info_WCR
 
country_code_categories<-unique(monkey_pox$country_code)
country_code_categories
 
 world_country_region<-world_country_region %>% 
  mutate(country_code= case_when(str_detect(alpha_2,"at")~"AT",
                          str_detect(alpha_2,"be")~"BE",
                          str_detect(alpha_2,"bg")~"BG",
                          str_detect(alpha_2,"cy")~"CY",
                          str_detect(alpha_2,"cz")~"CZ",
                          str_detect(alpha_2,"de")~"DE",
                          str_detect(alpha_2,"dk")~"DK",
                          str_detect(alpha_2,"ee")~"EE",
                          str_detect(alpha_2,"el")~"EL",
                          str_detect(alpha_2,"es")~"ES",
                          str_detect(alpha_2,"fi")~"FI",
                          str_detect(alpha_2,"fr")~"FR",
                          str_detect(alpha_2,"hr")~"HR",
                          str_detect(alpha_2,"hu")~"HU",
                          str_detect(alpha_2,"ie")~"IE",
                          str_detect(alpha_2,"is")~"IS",
                          str_detect(alpha_2,"it")~"IT",
                          str_detect(alpha_2,"lt")~"LT",
                          str_detect(alpha_2,"lu")~"LU",
                          str_detect(alpha_2,"lv")~"LV",
                          str_detect(alpha_2,"mt")~"MT",
                          str_detect(alpha_2,"nl")~"NL",
                          str_detect(alpha_2,"no")~"NO",
                          str_detect(alpha_2,"pl")~"PL",
                          str_detect(alpha_2,"pt")~"PT",
                          str_detect(alpha_2,"ro")~"RO",
                          str_detect(alpha_2,"se")~"SE",
                          str_detect(alpha_2,"si")~"SI",
                          str_detect(alpha_2,"sk")~"SK",
                         TRUE~NA_character_)) 
 
 world_country_region<- world_country_region %>% drop_na(country_code)
  
 
 world_country_region<- world_country_region%>%
   select(country_code, sub_region)

```



```{r,include=FALSE, echo=FALSE}
file_path3<-"https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_census_stats.csv"
census_stats <- read.csv(file_path3,na = c("", "NA", "*", "n/a"))%>% 
  clean_names()

var_info_CS <- data.frame(Variable = names(census_stats),
                      VariableType = sapply(census_stats, class),
                      MissingValues = sapply(census_stats, function(y) 
                        sum(length(which(is.na(y))))),
                      row.names = NULL)
var_info_CS

edu_categories<-unique(census_stats$edu)
edu_categories

cas_categories<-unique(census_stats$cas)
cas_categories

age_categories <- unique(census_stats$age)
age_categories
 
sex_categories<- unique(census_stats$sex)
sex_categories

census_stats<-census_stats %>% 
  mutate(edu= case_when(edu=="NONE" ~ "No formal education",
                        edu== "ED1" ~ "Primary education",
                        edu== "ED2" ~ "Lower secondary education",
                        edu== "ED3" ~ "Upper secondary education",
           edu== "ED4" ~ "Post secondary non-tertiary education (tradeschool)",
                  edu== "ED5" ~ "First stage of tertiary education (college)",
           edu== "ED6" ~ "Second stage of tertiary education (grad school)",
           TRUE~NA_character_))%>%
  drop_na(edu)%>%
  mutate(cas= case_when(cas== "ACT" ~ "Total economically active",
                        cas== "EMP" ~ "Employed (among economically active)",
                        cas== "UNE" ~ "Unemployed (among economically active)",
                        cas== "INAC" ~ "Total economically inactive",
                        TRUE~NA_character_))%>%
  drop_na(cas)%>%
  mutate(age=case_when(age== "Y_LT15" ~ " < 15",
                         age== "Y15-29" ~ "15-29",
                         age== "Y30-49" ~ "30-49",
                         age== "Y50-64" ~ "50-64",
                         age== "Y65-84" ~ "65-84",
                         TRUE ~ " 85+"))%>%
  mutate(sex=case_when(sex=="F"~ "Female",
                       TRUE~"Male"))

  census_stats_edu<- census_stats %>%
  group_by(country_code,edu) %>% 
  summarise(total_pop_edu = n())
  
  census_stats_edu <- census_stats_edu %>%
  pivot_wider(names_from = "edu", names_prefix = "edu_", 
              values_from = total_pop_edu)

census_stats_cas <- census_stats %>%
  group_by(country_code, cas)%>%
summarise(total_pop_cas = n())

census_stats_cas <- census_stats_cas%>%
  pivot_wider(names_from = "cas", names_prefix = "cas_", 
              values_from = total_pop_cas)

census_stats_sex <- census_stats %>%
  group_by(country_code, sex)%>%
summarise(total_pop_sex= n())
census_stats_sex <- census_stats_sex %>%
  pivot_wider(names_from = "sex", names_prefix = "sex_",
              values_from = total_pop_sex)

census_stats_age <- census_stats %>%
  group_by(country_code, age)%>%
summarise(total_pop_age= n())
census_stats_age<- census_stats_age %>%
  pivot_wider(names_from = "age", names_prefix = "age_", 
              values_from = total_pop_age)

census_stats_respop <- census_stats %>%
  group_by(country_code, res_pop)%>%
summarise(total_pop_respop= n())
census_stats_respop <- census_stats_respop%>%
  pivot_wider(names_from = "res_pop", names_prefix = "res_pop_",
              values_from = total_pop_respop)

                      

```



```{r, include=FALSE, echo=FALSE}
joined_df <- merge(monkey_pox, pop_denominator, by.x = "country_code", 
             by.y = "country_code", all.x = TRUE, all.y = FALSE)

var_info <- data.frame(Variable = names(joined_df),
                      VariableType = sapply(joined_df, class),
                      MissingValues = sapply(joined_df, function(y) 
                        sum(length(which(is.na(y))))),
                      row.names = NULL)
var_info

joined_df <- merge(joined_df, world_country_region, by.x = "country_code", 
             by.y = "country_code", all.x = TRUE, all.y = FALSE)

var_info <- data.frame(Variable = names(joined_df),
                      VariableType = sapply(joined_df, class),
                      MissingValues = sapply(joined_df, function(y) 
                        sum(length(which(is.na(y))))),
                      row.names = NULL)
var_info

joined_df <- joined_df %>% drop_na(sub_region)%>%
   filter(!(sub_region == "Western Asia"))

total_dataset<- joined_df%>% 
  group_by(country_code)%>%
  mutate(total_MPX_case = sum(total_conf_case))%>%
  select(country_code, total_MPX_case,time_period, sub_region)




total_dataset <- merge( total_dataset, census_stats_edu,by.x ="country_code", 
                       by.y = "country_code",all.x = TRUE, all.y = FALSE)

total_dataset <- merge( total_dataset, census_stats_cas,by.x ="country_code", 
                       by.y = "country_code",all.x = TRUE, all.y = FALSE)

total_dataset <- merge( total_dataset, census_stats_sex,by.x ="country_code", 
                       by.y = "country_code",all.x = TRUE, all.y = FALSE)


total_dataset <- merge( total_dataset, census_stats_age,by.x ="country_code", 
                       by.y = "country_code",all.x = TRUE, all.y = FALSE)

total_dataset<-merge( total_dataset,
                             census_stats_respop,by.x ="country_code", 
                       by.y = "country_code",all.x = TRUE, all.y = FALSE)

total_dataset<- total_dataset%>%
  filter(!duplicated(country_code))
total_dataset<- total_dataset%>% clean_names()
```

\newpage

**Methods**


**Data sources**

We used four data sets to tackle this project: Data on monkeypox cases in the
EU/EEA, European population denominators, European census statistics data 
and world country regions data. Data on monkeypox cases in the EU/EEA was 
accessed from the European Center for Disease Prevention and Control (ECDC).
The European population denominator data source is from the European commission,
 the EU census statistics data source is from European census statistics 2011.

**Data cleaning**

From data on monkeypox cases in the EU/EEA, we selected three variables named:
Country code, date cases were reported of which we generated months on which 
monkeypox cases were reported and the number of monkeypox cases reported.We then 
selected only two variables from population denominator data set being country 
code which was renamed from geo and time period, then we filtered the data set 
by time period of year 2022 in order for us to have the most recent time period. 

Further more we selected two variables from census data: country code and sex
because we were interested at the distribution of monkeypox rate by gender. 
Additionally , we selected two variables from world country region data set 
being country code and sub regions which were only for Europe by filtering only 
country code that matches the one that are available in monkeypox cases data set.

All the data sets were merged using country code to allow us generate new 
variables according to the objective of the project: we created monkeypox rate 
per region per month and monkeypox rate per sex for all countries in the Europe 
which reported monkey pox cases.The missing records and records that were not
applicable for our project were removed from the data set.


**Analytic Methods**

A table  and a plot containing bar chart and a line chart were created from our
data set after joining monkeypox cases data set, population denominator and 
world country region data set. The bar chart represents the monkeypox rate in
each european region categorised by different month when reported in year 2022,
while the line chart represents the monkeypox rate for different month in each 
region. Then the scatter plot was created to measure the linear relationship 
between monkeypox rate and sex.


\newpage



**Results** 



```{r, include=TRUE,warning=FALSE,message = F,echo=FALSE}
summary_table1<- joined_df%>%
  group_by(sub_region, month_rep)%>%
  summarise(total_case_region= sum(total_conf_case))%>%
  mutate(monthly_cases_per_region= sum(total_case_region), 
  MPX_cases_rate_per_region =
    round( total_case_region/monthly_cases_per_region*100,1))%>%
  arrange(sub_region, desc(month_rep))
datatable(summary_table1,rownames = FALSE ,
          caption = 'Table 1: Distribution of MonkeyPox Rate per month per region',
          colnames=c('Region','Month','MonkeyPox Cases',
                     'Total MonkeyPox', 'MonkeyPox Rate'))

```

Table 1 shows Monkeypox rates by regions from May to August in 2022.




\newpage



**Figure1**

```{r,include=TRUE,echo=FALSE}
                    
    
    summary_table1<-summary_table1 %>%
      mutate(date_months=case_when(month_rep=="May"~5,
                                   month_rep=="June"~6,
                                   month_rep=="July"~7,
                                   TRUE~8))

    
summary_table1$month_rep<- factor(summary_table1$month_rep) 
    summary_table1%>%
      ggplot(aes( x=sub_region, y= MPX_cases_rate_per_region,fill=month_rep)) +
      geom_bar(position="dodge",stat="identity")+
      labs(x = "EU Regions", y = "MonkeyPox Cases Rate")+
      ggtitle("Dodged bar chart of monkeypox case rates by 
      month for different regions within the EU")
    
 

```

Figure 1 shows a visualization of monkeypox case rates by month for different 
regions within the EU. There has been an increase in monkeypox case rates in 
all EU regions since May 2022, which then began to subside in August.

\newpage

**Figure2**


```{r,include=TRUE,warning=FALSE,echo=FALSE}

plot_ly(
  summary_table1,
  x=~ date_months,
  y=~MPX_cases_rate_per_region,
  color=~sub_region,
  type="scatter",
  mode="lines",
  colors=c("darkorange","darkcyan","darkslateblue", "darkred")
) %>%
  layout(
    title="MonkeyPox rate in Europe within 2022 for four months( May-August)",
    yaxis=list(title="MonkeyPox Rate"),
    xaxis=list(title="Month"),
    paper_bgcolor="azure",
    plot_bgcolor="white"
  )

```


Figure 2 shows the relationship of monkeypox rate and month for different region
of Europe.There has been an increase in monkeypox case rates in all EU regions 
since May 2022, which then began to subside in August.


\newpage

```{r,include=FALSE,echo=FALSE}
female_summary<- total_dataset%>%
  mutate(mpx_rate = round(total_mpx_case/ sum(sex_female) *100,2))%>%
  select(country_code, total_mpx_case, sex_female, mpx_rate)%>%
  rename(female_population = sex_female)
female_summary
```



**Figure3**

```{r,include=TRUE,echo=FALSE}
scatter_plot <- ggplot(data = female_summary, aes(x = female_population,
                                                 y = mpx_rate)) + 
  geom_point(na.rm=TRUE) + theme_minimal(base_size = 15)+
  labs(x = " Total Number of females", 
       y = "Total number of MonkeyPox cases",
  title ="The scatter plot of MonkeyPox cases and female population
                   within European counties")
scatter_plot
   
```


Figure 3 shows the scatter plot of between monkeypox rates and female population,
indicating the absence of linear relationship between our variables.


\newpage


```{r,include=FALSE,echo=FALSE}

male_summary<- total_dataset%>%
  mutate(mpx_rate = round(total_mpx_case/ sum(sex_male) *100,2))%>%
  select(country_code, total_mpx_case, sex_male, mpx_rate)%>%
  rename(male_population = sex_male)
female_summary

```

**Figure4**



```{r,include=TRUE,echo=FALSE}
scatter_plot <- ggplot(data = male_summary, aes(x = male_population,
                                                 y = mpx_rate)) + 
  geom_point(na.rm=TRUE) + theme_minimal(base_size = 15)+
  labs(x = " Total Number of males", 
       y = "Total number of MonkeyPox cases",
     title ="The scatter plot of MonkeyPox cases and male population
                   within European counties")
scatter_plot
```




Figure 4 shows the scatter plot of between monkeypox rates and male population,
indicating the absence of linear relationship between our variables.




\newpage

**Discussion**



The results of our analysis suggest that there is no correlation between gender
and monkeypox case rates, as can be seen in figure 3 and 4.
From figure  the monkeypox rates were less compared to other months, in July all 
the regions experience a huge increase of monkeypox rate of which Western Europe 
region had the highest rate compared to other regions. Figure 2 depicts that 
though case rates are increasing in all EU regions, there has been a drastic 
decrease in case rates during August especially for Southern Europe followed by
Western Europe. In the Eastern and Northern Europe, they are not subsiding as
quickly as in the other two regions. Based on our investigation, we suggest an
increase in surveillance in Eastern and Northern Europe as they need more support
in dealing with an outbreak in monkeypox cases.

