---
title: "Milestone_3"
author: "Moliehi Mokete and Bongekile Nkosi"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Loading the libraries**

```{r, include=T,echo=FALSE}
library(tidyverse)
library(readr)
library(janitor)
library(purrr)
library(stringr)
```

**loading the MPX data**
```{r}
file_path1<-"https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_mpx_cases.csv"
monkey_pox <-read_csv(file_path1,na=c("","NA","*", "n/a"),
                      show_col_types = FALSE)%>% clean_names()

monkey_pox<- monkey_pox%>% mutate(month_rep= months(date_rep))%>%
  group_by(country_code,month_rep)%>%
  mutate(total_conf_case = sum(conf_cases))

 var_info_MPX <- data.frame(Variable = names(monkey_pox),
                      VariableType = sapply(monkey_pox, class),
                      MissingValues = sapply(monkey_pox, function(y) 
                        sum(length(which(is.na(y))))),
                      row.names = NULL)
 var_info_MPX 
 
 monkey_pox <- monkey_pox %>%
arrange(country_code) %>%
group_by(month_rep) %>%
filter(!duplicated(country_code))
 
 monkey_pox<-monkey_pox %>% 
   select(country_code,month_rep,total_conf_case)

```
\newpage

**Loading Population denominator dataset**

```{r}
file_path2<-"https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_pop_denominators.csv" 
pop_denominator<- read.csv(file_path2,na = c("", "NA", "*", "n/a")) %>%
  clean_names() %>% 
  rename(country_code = geo)

 var_info_PD <- data.frame(Variable = names(pop_denominator),
                      VariableType = sapply(pop_denominator, class),
                      MissingValues = sapply(pop_denominator, function(y) 
                        sum(length(which(is.na(y))))),
                      row.names = NULL)
 var_info_PD 
 
 pop_denominator <- pop_denominator%>% 
   filter(time_period== 2022)%>%
   select(country_code, time_period)

```

\newpage

**Loading world country region dataset**

```{r}
file_path4<-"https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/world_country_regions.csv"
world_country_region <- read.csv(file_path4,na = c("", "NA", "*", "n/a"))%>%
  clean_names()

var_info_WCR <- data.frame(Variable = names(world_country_region),
                      VariableType = sapply(world_country_region, class),
                      MissingValues = sapply(world_country_region, function(y) 
                        sum(length(which(is.na(y))))),
                      row.names = NULL)
 var_info_WCR
 
country_code_categories<-unique(monkey_pox$country_code)
country_code_categories
 
 world_country_region<-world_country_region %>% 
  mutate(country_code= case_when(str_detect(alpha_2,"at")~"AT",
                          str_detect(alpha_2,"be")~"BE",
                          str_detect(alpha_2,"bg")~"BG",
                          str_detect(alpha_2,"cy")~"CY",
                          str_detect(alpha_2,"cz")~"CZ",
                          str_detect(alpha_2,"de")~"DE",
                          str_detect(alpha_2,"dk")~"DK",
                          str_detect(alpha_2,"ee")~"EE",
                          str_detect(alpha_2,"el")~"EL",
                          str_detect(alpha_2,"es")~"ES",
                          str_detect(alpha_2,"fi")~"FI",
                          str_detect(alpha_2,"fr")~"FR",
                          str_detect(alpha_2,"hr")~"HR",
                          str_detect(alpha_2,"hu")~"HU",
                          str_detect(alpha_2,"ie")~"IE",
                          str_detect(alpha_2,"is")~"IS",
                          str_detect(alpha_2,"it")~"IT",
                          str_detect(alpha_2,"lt")~"LT",
                          str_detect(alpha_2,"lu")~"lu",
                          str_detect(alpha_2,"lv")~"LV",
                          str_detect(alpha_2,"mt")~"MT",
                          str_detect(alpha_2,"nl")~"NL",
                          str_detect(alpha_2,"no")~"NO",
                          str_detect(alpha_2,"pl")~"PL",
                          str_detect(alpha_2,"pt")~"PT",
                          str_detect(alpha_2,"ro")~"RO",
                          str_detect(alpha_2,"se")~"SE",
                          str_detect(alpha_2,"si")~"SI",
                          str_detect(alpha_2,"sk")~"SK",
                         TRUE~NA_character_))%>%
   drop_na(country_code)
 
 world_country_region<- world_country_region%>%
   select(country_code, sub_region)

```
\newpage

**Loading Census Data set**

```{r}
file_path3<-"https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_census_stats.csv"
census_stats <- read.csv(file_path3,na = c("", "NA", "*", "n/a"))%>% 
  clean_names()

var_info_CS <- data.frame(Variable = names(census_stats),
                      VariableType = sapply(census_stats, class),
                      MissingValues = sapply(census_stats, function(y) 
                        sum(length(which(is.na(y))))),
                      row.names = NULL)
var_info_CS

edu_categories<-unique(census_stats$edu)
edu_categories

cas_categories<-unique(census_stats$cas)
cas_categories

age_categories <- unique(census_stats$age)
age_categories
 
sex_categories<- unique(census_stats$sex)
sex_categories

census_stats<-census_stats %>% 
  mutate(edu= case_when(edu=="NONE" ~ "No formal education",
                        edu== "ED1" ~ "Primary education",
                        edu== "ED2" ~ "Lower secondary education",
                        edu== "ED3" ~ "Upper secondary education",
           edu== "ED4" ~ "Post secondary non-tertiary education (tradeschool)",
                  edu== "ED5" ~ "First stage of tertiary education (college)",
           edu== "ED6" ~ "Second stage of tertiary education (grad school)",
           TRUE~NA_character_))%>%
  drop_na(edu)%>%
  mutate(cas= case_when(cas== "ACT" ~ "Total economically active",
                        cas== "EMP" ~ "Employed (among economically active)",
                        cas== "UNE" ~ "Unemployed (among economically active)",
                        cas== "INAC" ~ "Total economically inactive",
                        TRUE~NA_character_))%>%
  drop_na(cas)%>%
  mutate(age=case_when(age== "Y_LT15" ~ " < 15",
                         age== "Y15-29" ~ "15-29",
                         age== "Y30-49" ~ "30-49",
                         age== "Y50-64" ~ "50-64",
                         age== "Y65-84" ~ "65-84",
                         TRUE ~ " 85+"))%>%
  mutate(sex=case_when(sex=="F"~ "Female",
                       TRUE~"Male"))

  census_stats_edu<- census_stats %>%
  group_by(country_code,edu) %>% 
  summarise(total_pop_edu = n())

census_stats_cas <- census_stats %>%
  group_by(country_code, cas)%>%
summarise(total_pop_cas = n())

census_stats_sex <- census_stats %>%
  group_by(country_code, sex)%>%
summarise(total_pop_sex= n())
census_stats_age <- census_stats %>%
  group_by(country_code, age)%>%
summarise(total_pop_age= n())

census_stats_respop <- census_stats %>%
  group_by(country_code, res_pop)%>%
summarise(total_pop_respop= n())

                      

```

**Joining all data sets**

```{r}
joined_df <- merge(monkey_pox, pop_denominator, by.x = "country_code", 
             by.y = "country_code", all.x = TRUE, all.y = FALSE)

var_info <- data.frame(Variable = names(joined_df),
                      VariableType = sapply(joined_df, class),
                      MissingValues = sapply(joined_df, function(y) 
                        sum(length(which(is.na(y))))),
                      row.names = NULL)
var_info

joined_df <- merge(joined_df, world_country_region, by.x = "country_code", 
             by.y = "country_code", all.x = TRUE, all.y = FALSE)

var_info <- data.frame(Variable = names(joined_df),
                      VariableType = sapply(joined_df, class),
                      MissingValues = sapply(joined_df, function(y) 
                        sum(length(which(is.na(y))))),
                      row.names = NULL)
var_info

joined_df<- joined_df%>%
  mutate(sub_region=ifelse(country_code=="EL","Southeast Europe",sub_region),
  sub_region=ifelse(country_code=="LU","Northwestern Europe",sub_region))


joined_df_edu <- merge( joined_df, census_stats_edu,by.x ="country_code", 
                       by.y = "country_code",all.x = TRUE, all.y = FALSE)

joined_df_cas <- merge( joined_df, census_stats_cas,by.x ="country_code", 
                       by.y = "country_code",all.x = TRUE, all.y = FALSE)

joined_df_sex <- merge( joined_df, census_stats_sex,by.x ="country_code", 
                       by.y = "country_code",all.x = TRUE, all.y = FALSE)


joined_df_age <- merge( joined_df, census_stats_age,by.x ="country_code", 
                       by.y = "country_code",all.x = TRUE, all.y = FALSE)

joined_df_popdensity<- merge( joined_df, census_stats_respop,by.x ="country_code", 
                       by.y = "country_code",all.x = TRUE, all.y = FALSE)

```


\newpage
Data dictionary based on clean dataset (minimum 4 data elements), including:
Variable name
Data type
Description

```{r}
data_dict <- function(joined_df, desc = c()){
  data.frame(
    "Variable Name" = names(joined_df),
    "Variable Type" = sapply(joined_df,class),
    "Variable Description" = desc, 
    check.names = FALSE, row.names = NULL
  )
}

data_dict(joined_df[], desc =c(
  "country code ", 
  "months cases were reported",
  "total MPX cases recorded",
  " the recent time period",
  "countries sub regions in Europe"))

data_dict <- function(joined_df_age, desc = c()){
  data.frame(
    "Variable Name" = names(joined_df_age),
    "Variable Type" = sapply(joined_df_age,class),
    "Variable Description" = desc, 
    check.names = FALSE, row.names = NULL
  )
}

data_dict(joined_df_age[],desc=c(
   "country code ", 
  "months cases were reported",
  "total MPX cases recorded",
  " the recent time period",
  "countries sub regions in Europe",
  " age groups of the population",
  "total population per age group"))

data_dict <- function(joined_df_cas, desc = c()){
  data.frame(
    "Variable Name" = names(joined_df_cas),
    "Variable Type" = sapply(joined_df_cas,class),
    "Variable Description" = desc, 
    check.names = FALSE, row.names = NULL
  )
}

data_dict(joined_df_cas[],desc=c(
  "country code ", 
  "months cases were reported",
  "total MPX cases recorded",
  "the recent time period",
  "countries sub regions in Europe",
  "economical status of the population",
  "total population per economical status"))

data_dict <- function(joined_df_edu, desc = c()){
  data.frame(
    "Variable Name" = names(joined_df_edu),
    "Variable Type" = sapply(joined_df_edu,class),
    "Variable Description" = desc, 
    check.names = FALSE, row.names = NULL
  )
}

data_dict(joined_df_edu[],desc=c(
   "country code ", 
  "months cases were reported",
  "total MPX cases recorded",
  "the recent time period",
  "countries sub regions in Europe",
  "the categories of education level",
  "total population per education level"))

data_dict <- function(joined_df_sex, desc = c()){
  data.frame(
    "Variable Name" = names(joined_df_sex),
    "Variable Type" = sapply(joined_df_sex,class),
    "Variable Description" = desc, 
    check.names = FALSE, row.names = NULL
  )
}

data_dict(joined_df_sex[],desc=c(
  "country code ", 
  "months cases were reported",
  "total MPX cases recorded",
  " the recent time period",
  "countries sub regions in Europe",
  " sex of the population",
  " total population per sex"))

data_dict <- function(joined_df_popdensity, desc = c()){
  data.frame(
    "Variable Name" = names(joined_df_popdensity),
    "Variable Type" = sapply(joined_df_popdensity,class),
    "Variable Description" = desc, 
    check.names = FALSE, row.names = NULL
  )
}

data_dict(joined_df_popdensity [],desc=c(
  "country code ", 
  "months cases were reported",
  "total MPX cases recorded",
  " the recent time period",
  "countries sub regions in Europe", 
  "categories of population density",
  "total population per population density"))
```

\newpage
One or more tables with descriptive statistics for 4 data element

```{r}
summary_stas<-joined_df%>%
  group_by(month_rep)%>%
  summarise(mean= mean(total_conf_case), sd=sd(total_conf_case),
            max=max(total_conf_case),min=min(total_conf_case))
summary_stas

summary_stas<-joined_df%>%
  group_by(sub_region)%>%
  summarise(mean_cases= mean(total_conf_case), sd_cases=sd(total_conf_case),
            max_cases=max(total_conf_case),min_cases=min(total_conf_case))
summary_stas
summary_stas<-joined_df_edu%>%
  group_by(sub_region)%>%
  summarise(mean_edu= mean(total_pop_edu), sd_edu=sd(total_pop_edu),
            max_edu=max(total_pop_edu),min_edu=min(total_pop_edu))
summary_stas

summary_stas<-joined_df_cas%>%
  group_by(sub_region)%>%
  summarise(mean_cas= mean(total_pop_cas), sd_cas=sd(total_pop_cas),
            max_cas=max(total_pop_cas),min_cas=min(total_pop_cas))
summary_stas

summary_stas<-joined_df_age%>%
  group_by(sub_region)%>%
  summarise(mean_age= mean(total_pop_age), sd_age=sd(total_pop_age),
            max_age=max(total_pop_age),min_age=min(total_pop_age))
summary_stas

  
```

