---
title: "Milestone_6"
author: "Moliehi Mokete and Bongekile Nkosi"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Problem statement**

With the rise of MonkeyPox cases in Europe, in preparation of an appropriate
response, our team has been tasked with developing an update for the leadership 
of our state health department regarding the state of MPX. We are specifically 
interested in understanding how case rates may differ by region and various 
demographic factors. The demographic factor we are particularly interested in
is gender.

```{r, include=FALSE,echo=FALSE}
library(tidyverse)
library(readr)
library(janitor)
library(purrr)
library(stringr)
library(kableExtra)
library(plotly)
library(DT)
library(lubridate)
```


```{r,include=FALSE, echo=FALSE}
file_path1<-"https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_mpx_cases.csv"
monkey_pox <-read_csv(file_path1,na=c("","NA","*", "n/a"),
                      show_col_types = FALSE)%>% clean_names()

monkey_pox<- monkey_pox%>% mutate(month_rep= months(date_rep))%>%
  group_by(country_code,month_rep)%>%
  summarise(total_conf_case = sum(conf_cases))



 var_info_MPX <- data.frame(Variable = names(monkey_pox),
                      VariableType = sapply(monkey_pox, class),
                      MissingValues = sapply(monkey_pox, function(y) 
                        sum(length(which(is.na(y))))),
                      row.names = NULL)
 var_info_MPX 
 
 
 monkey_pox<-monkey_pox %>% 
   select(country_code,month_rep,total_conf_case)

```




```{r,include=FALSE,echo=FALSE}
file_path2<-"https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_pop_denominators.csv" 
pop_denominator<- read.csv(file_path2,na = c("", "NA", "*", "n/a")) %>%
  clean_names() %>% 
  rename(country_code = geo)

 var_info_PD <- data.frame(Variable = names(pop_denominator),
                      VariableType = sapply(pop_denominator, class),
                      MissingValues = sapply(pop_denominator, function(y) 
                        sum(length(which(is.na(y))))),
                      row.names = NULL)
 var_info_PD 
 
 pop_denominator <- pop_denominator%>% 
   filter(time_period== 2022)%>%
   select(country_code, time_period)

```


```{r,include=FALSE,echo=FALSE}
file_path4<-"https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/world_country_regions.csv"
world_country_region <- read.csv(file_path4,na = c("", "NA", "*", "n/a"))%>%
  clean_names()

var_info_WCR <- data.frame(Variable = names(world_country_region),
                      VariableType = sapply(world_country_region, class),
                      MissingValues = sapply(world_country_region, function(y) 
                        sum(length(which(is.na(y))))),
                      row.names = NULL)
 var_info_WCR
 
country_code_categories<-unique(monkey_pox$country_code)
country_code_categories
 
 world_country_region<-world_country_region %>% 
  mutate(country_code= case_when(str_detect(alpha_2,"at")~"AT",
                          str_detect(alpha_2,"be")~"BE",
                          str_detect(alpha_2,"bg")~"BG",
                          str_detect(alpha_2,"cy")~"CY",
                          str_detect(alpha_2,"cz")~"CZ",
                          str_detect(alpha_2,"de")~"DE",
                          str_detect(alpha_2,"dk")~"DK",
                          str_detect(alpha_2,"ee")~"EE",
                          str_detect(alpha_2,"el")~"EL",
                          str_detect(alpha_2,"es")~"ES",
                          str_detect(alpha_2,"fi")~"FI",
                          str_detect(alpha_2,"fr")~"FR",
                          str_detect(alpha_2,"hr")~"HR",
                          str_detect(alpha_2,"hu")~"HU",
                          str_detect(alpha_2,"ie")~"IE",
                          str_detect(alpha_2,"is")~"IS",
                          str_detect(alpha_2,"it")~"IT",
                          str_detect(alpha_2,"lt")~"LT",
                          str_detect(alpha_2,"lu")~"LU",
                          str_detect(alpha_2,"lv")~"LV",
                          str_detect(alpha_2,"mt")~"MT",
                          str_detect(alpha_2,"nl")~"NL",
                          str_detect(alpha_2,"no")~"NO",
                          str_detect(alpha_2,"pl")~"PL",
                          str_detect(alpha_2,"pt")~"PT",
                          str_detect(alpha_2,"ro")~"RO",
                          str_detect(alpha_2,"se")~"SE",
                          str_detect(alpha_2,"si")~"SI",
                          str_detect(alpha_2,"sk")~"SK",
                         TRUE~NA_character_)) 
 
 world_country_region<- world_country_region %>% drop_na(country_code)
  
 
 world_country_region<- world_country_region%>%
   select(country_code, sub_region)

```



```{r,include=FALSE, echo=FALSE}
file_path3<-"https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_census_stats.csv"
census_stats <- read.csv(file_path3,na = c("", "NA", "*", "n/a"))%>% 
  clean_names()

var_info_CS <- data.frame(Variable = names(census_stats),
                      VariableType = sapply(census_stats, class),
                      MissingValues = sapply(census_stats, function(y) 
                        sum(length(which(is.na(y))))),
                      row.names = NULL)
var_info_CS

edu_categories<-unique(census_stats$edu)
edu_categories

cas_categories<-unique(census_stats$cas)
cas_categories

age_categories <- unique(census_stats$age)
age_categories
 
sex_categories<- unique(census_stats$sex)
sex_categories

census_stats<-census_stats %>% 
  mutate(edu= case_when(edu=="NONE" ~ "No formal education",
                        edu== "ED1" ~ "Primary education",
                        edu== "ED2" ~ "Lower secondary education",
                        edu== "ED3" ~ "Upper secondary education",
           edu== "ED4" ~ "Post secondary non-tertiary education (tradeschool)",
                  edu== "ED5" ~ "First stage of tertiary education (college)",
           edu== "ED6" ~ "Second stage of tertiary education (grad school)",
           TRUE~NA_character_))%>%
  drop_na(edu)%>%
  mutate(cas= case_when(cas== "ACT" ~ "Total economically active",
                        cas== "EMP" ~ "Employed (among economically active)",
                        cas== "UNE" ~ "Unemployed (among economically active)",
                        cas== "INAC" ~ "Total economically inactive",
                        TRUE~NA_character_))%>%
  drop_na(cas)%>%
  mutate(age=case_when(age== "Y_LT15" ~ " < 15",
                         age== "Y15-29" ~ "15-29",
                         age== "Y30-49" ~ "30-49",
                         age== "Y50-64" ~ "50-64",
                         age== "Y65-84" ~ "65-84",
                         TRUE ~ " 85+"))%>%
  mutate(sex=case_when(sex=="F"~ "Female",
                       TRUE~"Male"))

  census_stats_edu<- census_stats %>%
  group_by(country_code,edu) %>% 
  summarise(total_pop_edu = n())
  
  census_stats_edu <- census_stats_edu %>%
  pivot_wider(names_from = "edu", names_prefix = "edu_", 
              values_from = total_pop_edu)

census_stats_cas <- census_stats %>%
  group_by(country_code, cas)%>%
summarise(total_pop_cas = n())

census_stats_cas <- census_stats_cas%>%
  pivot_wider(names_from = "cas", names_prefix = "cas_", 
              values_from = total_pop_cas)

census_stats_sex <- census_stats %>%
  group_by(country_code, sex)%>%
summarise(total_pop_sex= n())
census_stats_sex <- census_stats_sex %>%
  pivot_wider(names_from = "sex", names_prefix = "sex_",
              values_from = total_pop_sex)

census_stats_age <- census_stats %>%
  group_by(country_code, age)%>%
summarise(total_pop_age= n())
census_stats_age<- census_stats_age %>%
  pivot_wider(names_from = "age", names_prefix = "age_", 
              values_from = total_pop_age)

census_stats_respop <- census_stats %>%
  group_by(country_code, res_pop)%>%
summarise(total_pop_respop= n())
census_stats_respop <- census_stats_respop%>%
  pivot_wider(names_from = "res_pop", names_prefix = "res_pop_",
              values_from = total_pop_respop)

                      

```



```{r, include=FALSE, echo=FALSE}
joined_df <- merge(monkey_pox, pop_denominator, by.x = "country_code", 
             by.y = "country_code", all.x = TRUE, all.y = FALSE)

var_info <- data.frame(Variable = names(joined_df),
                      VariableType = sapply(joined_df, class),
                      MissingValues = sapply(joined_df, function(y) 
                        sum(length(which(is.na(y))))),
                      row.names = NULL)
var_info

joined_df <- merge(joined_df, world_country_region, by.x = "country_code", 
             by.y = "country_code", all.x = TRUE, all.y = FALSE)

var_info <- data.frame(Variable = names(joined_df),
                      VariableType = sapply(joined_df, class),
                      MissingValues = sapply(joined_df, function(y) 
                        sum(length(which(is.na(y))))),
                      row.names = NULL)
var_info

joined_df <- joined_df %>% drop_na(sub_region)%>%
   filter(!(sub_region == "Western Asia"))

total_dataset<- joined_df%>% 
  group_by(country_code)%>%
  mutate(total_MPX_case = sum(total_conf_case))


total_dataset1 <- merge( total_dataset, census_stats_sex,by.x ="country_code", 
                       by.y = "country_code",all.x = TRUE, all.y = FALSE)

total_dataset <- merge( total_dataset, census_stats_edu,by.x ="country_code", 
                       by.y = "country_code",all.x = TRUE, all.y = FALSE)

total_dataset <- merge( total_dataset, census_stats_cas,by.x ="country_code", 
                       by.y = "country_code",all.x = TRUE, all.y = FALSE)




total_dataset <- merge( total_dataset, census_stats_age,by.x ="country_code", 
                       by.y = "country_code",all.x = TRUE, all.y = FALSE)

total_dataset<-merge( total_dataset,
                             census_stats_respop,by.x ="country_code", 
                       by.y = "country_code",all.x = TRUE, all.y = FALSE)

total_dataset<- total_dataset%>%
  filter(!duplicated(country_code))
total_dataset<- total_dataset%>% clean_names()

```

\newpage

**Methods**


**Data sources**

We used four data sets to tackle this project: Data on monkeypox cases in the
EU/EEA, European population denominators, European census statistics data 
and world country regions data. Data on monkeypox cases in the EU/EEA was 
accessed from the European Center for Disease Prevention and Control (ECDC).
The European population denominator data source is from the European commission,
 the EU census statistics data source is from European census statistics 2011.

**Data cleaning**

From data on monkeypox cases in the EU/EEA, we selected three variables named:
Country code, date cases were reported of which we generated months on which 
monkeypox cases were reported and the number of monkeypox cases reported.We then 
selected only two variables from population denominator data set being country 
code which was renamed from geo and time period, then we filtered the data set 
by time period of year 2022 in order for us to have the most recent time period. 

Further more we selected two variables from census data: country code and sex
because we were interested at the distribution of monkeypox rate by gender. 
Additionally , we selected two variables from world country region data set 
being country code and sub regions which were only for Europe by filtering only 
country code that matches the one that are available in monkeypox cases data set.

All the data sets were merged using country code to allow us generate new 
variables according to the objective of the project: we created monkeypox rate 
per region per month.The missing records and records that were not
applicable for our project were removed from the data set. The population variable
was generated by adding both female population and male population, and created
sex ratio to determine the distribution of sex in each country.


**Analytic Methods**

A table  and a plot containing bar chart  were created from our data set after
joining monkeypox cases data set, population denominator and 
world country region data set. The bar chart represents the monkeypox rate in
each European region categorised by different month when reported in year 2022.



\newpage



**Results** 



```{r, include=TRUE,warning=FALSE,message = F,echo=FALSE}

total_dataset1<- total_dataset1%>% 
  mutate(total_population = sex_Female+sex_Male)
summary_table1<- total_dataset1%>%
  group_by(sub_region, month_rep)%>%
  summarise(total_case_region= sum(total_conf_case),
            total_population = sum(total_population))%>%
 mutate(
 MPX_cases_rate_per_region =
  round( total_case_region/total_population*100,1))%>%
  arrange( desc(month_rep))
datatable(summary_table1,rownames = FALSE ,
          caption = 'Table 1: Distribution of MonkeyPox Rate per month per region',
          colnames=c('Region','Month','MonkeyPox Cases',
                     'Total population', 'MonkeyPox Rate'))

```

Table 1 shows Monkeypox rates by regions from May to August in 2022.




\newpage



**Figure1**

```{r,include=TRUE,echo=FALSE}
                    
  

    
summary_table1$month_rep<- factor(summary_table1$month_rep) 
    summary_table1%>%
      ggplot(aes( x= month_rep,y= MPX_cases_rate_per_region,fill=sub_region)) +
      geom_bar(position="dodge",stat="identity")+
      labs(x = "Months cases were reported", y = "MonkeyPox Cases Rate")+
      ggtitle("Distribution of monkeypox case rates by month for different 
regions within the EU")
    
 

```

Figure 1 shows a visualization of monkeypox case rates by month for different 
regions within the EU. We have explored the distribution of rates from recent 
month to old. Comparing August to July, rates were less in August compared to
July. July has most highest rate recorded among other months.
There has been an increase in monkeypox case rates in all EU regions since May 
2022, which then began to subside in August.

\newpage


**Discussion**


From figure 1 the monkeypox rates were less in May compared to other months, 
in July all the regions experience a huge increase of monkeypox rate of which
Western Europe region had the highest rate compared to other regions. Southern 
and Western Europe regions recorded highest monkey pox cases from May to August.


Based on our investigation, we suggest an increase in surveillance in Western 
and Southern Europe as they need more support in dealing with an outbreak in
monkeypox cases which means that the department of health in Europe should 
implement the strategies aimed at minimizing the infection of Monkeypox. 
July experience a peak of reported cases, which suggests that during hot seasons
the cases increases compared to cold seasons, the department of health can 
allocate the funds meant for assistance of monkeypox during hot seasons.

