---
title: "Milestone_4"
author: "Moliehi Mokete and Bongekile Nkosi"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r, include=FALSE,echo=FALSE}
library(tidyverse)
library(readr)
library(janitor)
library(purrr)
library(stringr)
library(kableExtra)
library(plotly)
```


```{r,include=FALSE, echo=FALSE}
file_path1<-"https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_mpx_cases.csv"
monkey_pox <-read_csv(file_path1,na=c("","NA","*", "n/a"),
                      show_col_types = FALSE)%>% clean_names()

monkey_pox<- monkey_pox%>% mutate(month_rep= months(date_rep))%>%
  group_by(country_code,month_rep)%>%
  summarise(total_conf_case = sum(conf_cases))



 var_info_MPX <- data.frame(Variable = names(monkey_pox),
                      VariableType = sapply(monkey_pox, class),
                      MissingValues = sapply(monkey_pox, function(y) 
                        sum(length(which(is.na(y))))),
                      row.names = NULL)
 var_info_MPX 
 
 
 monkey_pox<-monkey_pox %>% 
   select(country_code,month_rep,total_conf_case)

```




```{r,include=FALSE,echo=FALSE}
file_path2<-"https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_pop_denominators.csv" 
pop_denominator<- read.csv(file_path2,na = c("", "NA", "*", "n/a")) %>%
  clean_names() %>% 
  rename(country_code = geo)

 var_info_PD <- data.frame(Variable = names(pop_denominator),
                      VariableType = sapply(pop_denominator, class),
                      MissingValues = sapply(pop_denominator, function(y) 
                        sum(length(which(is.na(y))))),
                      row.names = NULL)
 var_info_PD 
 
 pop_denominator <- pop_denominator%>% 
   filter(time_period== 2022)%>%
   select(country_code, time_period)

```


```{r,include=FALSE,echo=FALSE}
file_path4<-"https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/world_country_regions.csv"
world_country_region <- read.csv(file_path4,na = c("", "NA", "*", "n/a"))%>%
  clean_names()

var_info_WCR <- data.frame(Variable = names(world_country_region),
                      VariableType = sapply(world_country_region, class),
                      MissingValues = sapply(world_country_region, function(y) 
                        sum(length(which(is.na(y))))),
                      row.names = NULL)
 var_info_WCR
 
country_code_categories<-unique(monkey_pox$country_code)
country_code_categories
 
 world_country_region<-world_country_region %>% 
  mutate(country_code= case_when(str_detect(alpha_2,"at")~"AT",
                          str_detect(alpha_2,"be")~"BE",
                          str_detect(alpha_2,"bg")~"BG",
                          str_detect(alpha_2,"cy")~"CY",
                          str_detect(alpha_2,"cz")~"CZ",
                          str_detect(alpha_2,"de")~"DE",
                          str_detect(alpha_2,"dk")~"DK",
                          str_detect(alpha_2,"ee")~"EE",
                          str_detect(alpha_2,"el")~"EL",
                          str_detect(alpha_2,"es")~"ES",
                          str_detect(alpha_2,"fi")~"FI",
                          str_detect(alpha_2,"fr")~"FR",
                          str_detect(alpha_2,"hr")~"HR",
                          str_detect(alpha_2,"hu")~"HU",
                          str_detect(alpha_2,"ie")~"IE",
                          str_detect(alpha_2,"is")~"IS",
                          str_detect(alpha_2,"it")~"IT",
                          str_detect(alpha_2,"lt")~"LT",
                          str_detect(alpha_2,"lu")~"LU",
                          str_detect(alpha_2,"lv")~"LV",
                          str_detect(alpha_2,"mt")~"MT",
                          str_detect(alpha_2,"nl")~"NL",
                          str_detect(alpha_2,"no")~"NO",
                          str_detect(alpha_2,"pl")~"PL",
                          str_detect(alpha_2,"pt")~"PT",
                          str_detect(alpha_2,"ro")~"RO",
                          str_detect(alpha_2,"se")~"SE",
                          str_detect(alpha_2,"si")~"SI",
                          str_detect(alpha_2,"sk")~"SK",
                         TRUE~NA_character_)) 
 
 world_country_region<- world_country_region %>% drop_na(country_code)
  
 
 world_country_region<- world_country_region%>%
   select(country_code, sub_region)

```



```{r,include=FALSE, echo=FALSE}
file_path3<-"https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/euro_census_stats.csv"
census_stats <- read.csv(file_path3,na = c("", "NA", "*", "n/a"))%>% 
  clean_names()

var_info_CS <- data.frame(Variable = names(census_stats),
                      VariableType = sapply(census_stats, class),
                      MissingValues = sapply(census_stats, function(y) 
                        sum(length(which(is.na(y))))),
                      row.names = NULL)
var_info_CS

edu_categories<-unique(census_stats$edu)
edu_categories

cas_categories<-unique(census_stats$cas)
cas_categories

age_categories <- unique(census_stats$age)
age_categories
 
sex_categories<- unique(census_stats$sex)
sex_categories

census_stats<-census_stats %>% 
  mutate(edu= case_when(edu=="NONE" ~ "No formal education",
                        edu== "ED1" ~ "Primary education",
                        edu== "ED2" ~ "Lower secondary education",
                        edu== "ED3" ~ "Upper secondary education",
           edu== "ED4" ~ "Post secondary non-tertiary education (tradeschool)",
                  edu== "ED5" ~ "First stage of tertiary education (college)",
           edu== "ED6" ~ "Second stage of tertiary education (grad school)",
           TRUE~NA_character_))%>%
  drop_na(edu)%>%
  mutate(cas= case_when(cas== "ACT" ~ "Total economically active",
                        cas== "EMP" ~ "Employed (among economically active)",
                        cas== "UNE" ~ "Unemployed (among economically active)",
                        cas== "INAC" ~ "Total economically inactive",
                        TRUE~NA_character_))%>%
  drop_na(cas)%>%
  mutate(age=case_when(age== "Y_LT15" ~ " < 15",
                         age== "Y15-29" ~ "15-29",
                         age== "Y30-49" ~ "30-49",
                         age== "Y50-64" ~ "50-64",
                         age== "Y65-84" ~ "65-84",
                         TRUE ~ " 85+"))%>%
  mutate(sex=case_when(sex=="F"~ "Female",
                       TRUE~"Male"))

  census_stats_edu<- census_stats %>%
  group_by(country_code,edu) %>% 
  summarise(total_pop_edu = n())
  
  census_stats_edu <- census_stats_edu %>%
  pivot_wider(names_from = "edu", names_prefix = "edu_", 
              values_from = total_pop_edu)

census_stats_cas <- census_stats %>%
  group_by(country_code, cas)%>%
summarise(total_pop_cas = n())

census_stats_cas <- census_stats_cas%>%
  pivot_wider(names_from = "cas", names_prefix = "cas_", 
              values_from = total_pop_cas)

census_stats_sex <- census_stats %>%
  group_by(country_code, sex)%>%
summarise(total_pop_sex= n())
census_stats_sex <- census_stats_sex %>%
  pivot_wider(names_from = "sex", names_prefix = "sex_",
              values_from = total_pop_sex)

census_stats_age <- census_stats %>%
  group_by(country_code, age)%>%
summarise(total_pop_age= n())
census_stats_age<- census_stats_age %>%
  pivot_wider(names_from = "age", names_prefix = "age_", 
              values_from = total_pop_age)

census_stats_respop <- census_stats %>%
  group_by(country_code, res_pop)%>%
summarise(total_pop_respop= n())
census_stats_respop <- census_stats_respop%>%
  pivot_wider(names_from = "res_pop", names_prefix = "res_pop_",
              values_from = total_pop_respop)

                      

```



```{r, include=FALSE, echo=FALSE}
joined_df <- merge(monkey_pox, pop_denominator, by.x = "country_code", 
             by.y = "country_code", all.x = TRUE, all.y = FALSE)

var_info <- data.frame(Variable = names(joined_df),
                      VariableType = sapply(joined_df, class),
                      MissingValues = sapply(joined_df, function(y) 
                        sum(length(which(is.na(y))))),
                      row.names = NULL)
var_info

joined_df <- merge(joined_df, world_country_region, by.x = "country_code", 
             by.y = "country_code", all.x = TRUE, all.y = FALSE)

var_info <- data.frame(Variable = names(joined_df),
                      VariableType = sapply(joined_df, class),
                      MissingValues = sapply(joined_df, function(y) 
                        sum(length(which(is.na(y))))),
                      row.names = NULL)
var_info

joined_df <- joined_df %>% drop_na(sub_region)%>%
   filter(!(sub_region == "Western Asia"))

total_dataset<- joined_df%>% 
  group_by(country_code)%>%
  mutate(total_MPX_case = sum(total_conf_case))%>%
  select(country_code, total_MPX_case,time_period, sub_region)




total_dataset <- merge( total_dataset, census_stats_edu,by.x ="country_code", 
                       by.y = "country_code",all.x = TRUE, all.y = FALSE)

total_dataset <- merge( total_dataset, census_stats_cas,by.x ="country_code", 
                       by.y = "country_code",all.x = TRUE, all.y = FALSE)

total_dataset <- merge( total_dataset, census_stats_sex,by.x ="country_code", 
                       by.y = "country_code",all.x = TRUE, all.y = FALSE)


total_dataset <- merge( total_dataset, census_stats_age,by.x ="country_code", 
                       by.y = "country_code",all.x = TRUE, all.y = FALSE)

total_dataset<-merge( total_dataset,
                             census_stats_respop,by.x ="country_code", 
                       by.y = "country_code",all.x = TRUE, all.y = FALSE)

total_dataset<- total_dataset%>%
  filter(!duplicated(country_code))
total_dataset<- total_dataset%>% clean_names()
```

\newpage
**1. Quality table as requested in scenario:**
**Case rates by month for different regions within the EU.**


```{r}
summary_table1<- joined_df%>%
  group_by(sub_region, month_rep)%>%
  summarise(total_case_region= sum(total_conf_case))%>%
  mutate(monthly_cases_per_region= sum(total_case_region), 
  MPX_cases_rate_per_region =
    round( total_case_region/monthly_cases_per_region*100,1))%>%
  arrange(sub_region, desc(month_rep))

summary_table1

```
The table 1 displays the distribution of Monkey Pox cases rate per month within 
different regions of Europe in 2022.


\newpage



**1.1  Quality plot or chart as requested in scenario**


```{r}
                    
    
    summary_table1<-summary_table1 %>%
      mutate(date_months=case_when(month_rep=="May"~5,
                                   month_rep=="June"~6,
                                   month_rep=="July"~7,
                                   TRUE~8))%>% arrange(date_months)
    
summary_table1$month_rep<- factor(summary_table1$month_rep) 
    summary_table1%>%
      ggplot(aes( x=sub_region, y= MPX_cases_rate_per_region,fill=month_rep)) +
      geom_bar(position="dodge",stat="identity")+
      labs(x = "EU Regions", y = "MonkeyPox Cases Rate")+
      ggtitle("Dodged bar chart of monkeypox case rates by 
      month for different regions within the EU")
    
 

```


**Interpretation:**


The dodged bar chart describes the distribution of monkey pox cases rate by
month among four regions of Europe. Among four regions of Europe, Western Europe
has the most reported monkey pox cases rate during the month of July. In July 
monkey pox cases rate were mostly reported compared to other months. The month
of May is the month where the cases of monkey pox in all regions of Europe were 
least reported, the Southern Europe reported more cases in May compared to other
European regions.


\newpage

**1.2  Quality plot or chart as requested in scenario**


```{r}

plot_ly(
  summary_table1,
  x=~ date_months,
  y=~MPX_cases_rate_per_region,
  color=~sub_region,
  type="scatter",
  mode="lines",
  colors=c("darkorange","darkcyan","darkslateblue", "darkred")
) %>%
  layout(
    title="MonkeyPox rate in Europe within 2022 for four months( May-August)",
    yaxis=list(title="MonkeyPox Rate"),
    xaxis=list(title="Month"),
    paper_bgcolor="azure",
    plot_bgcolor="white"
  )

```


**Interpretation:**


The scatter plot depicts the relationship of monkey pox cases rate and months 
cases were reported. The plot shows the increasing cases from May to July with 
Western Europe region have the highest increase of monkey pox rates compared to 
other European region. Then there is a decrease of monkey pox cases rate  from 
July to August , the Southern Europe region experienced most decrease as 
compared to other European regions. During the month of July , all the European
regions reported more cases rate compared to other months and Western Europe 
region reported most cases rate during the month of July. Through out the months 
of reporting, Northern and Eastern Europe regions were reporting the monkey pox 
cases rate which is more likely the same .


\newpage

```{r}
female_summary<- total_dataset%>%
  mutate(mpx_rate = round(total_mpx_case/ sum(sex_female) *100,2))%>%
  select(country_code, total_mpx_case, sex_female, mpx_rate)%>%
  rename(female_population = sex_female)
female_summary
```


The table 2 shows the distribution of monkey pox cases rate per female population
in each county.



```{r}
scatter_plot <- ggplot(data = female_summary, aes(x = female_population,
                                                 y = mpx_rate)) + 
  geom_point(na.rm=TRUE) + theme_minimal(base_size = 15)+
  labs(x = " Total Number of females", 
       y = "Total number of MonkeyPox cases",
  title ="The scatter plot of MonkeyPox cases and female population
                   within European counties")
scatter_plot
   
```


**Interpretation:**


The scatter plot shows the relationship between monkey pox cases rate and female
population. Looking at the plot there is no relationship between monkey pox cases 
rate and female population, that is, the patterns of the dots in the plot does not 
inform any relationship between the two variables. Therefore there is nonlinear 
relationship between female population and monkey pox cases rate in Europe.



```{r}

male_summary<- total_dataset%>%
  mutate(mpx_rate = round(total_mpx_case/ sum(sex_male) *100,2))%>%
  select(country_code, total_mpx_case, sex_male, mpx_rate)%>%
  rename(male_population = sex_male)
female_summary

```

The table 3 shows the distribution of monkey pox cases rate per male population
in each county.



```{r}
scatter_plot <- ggplot(data = male_summary, aes(x = male_population,
                                                 y = mpx_rate)) + 
  geom_point(na.rm=TRUE) + theme_minimal(base_size = 15)+
  labs(x = " Total Number of males", 
       y = "Total number of MonkeyPox cases",
     title ="The scatter plot of MonkeyPox cases and male population
                   within European counties")
scatter_plot
```



**Interpretation:**


The scatter plot shows the relationship between monkey pox cases rate and male
population. Looking at the plot there is no relationship between monkey pox cases 
rate and male population, that is, the patterns of the dots in the plot does not 
inform any relationship between the two variables.Therefore there is nonlinear 
relationship between male population and monkey pox cases rate in Europe.



